### 2017-06-08 Manuscripts/cog-psych/cogpsych.tex
framing applied to remainder of document

\nbcnote{More strongly link this to creativity: what do these findings indicate?}
\jlanote{ok with me now.}

@@

\subsection{Something Different: A Role For Contrast}

@@

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, it is assumed that learners in previous experiments were {\em successful} in creating new categories. Thus, effects of category contrast should be observable within the experimental results of these studies. To provide a test of the influence of category contrast in creative generation, we conducted a novel analysis of Experiment 3 from \cite{jern2013probabilistic}.

\nbcnote{Does it feel strange to focus on Experiment 3 specifically? Do you think wee need to explain why this is the `right' experiment for our question?}

\jlanote{I'd add a footnote explaining why we focus on this one only}

@@

Although we cannot know precisely which colors were displayed or perceived, we can still analyze their results from a coarse perspective to see whether there is preliminary support for contrast. To do so, we binned all possible hues into one of eight uniformly-spaced color groups: \{{\em Red}: $0-0.063$, $0.938-1$, {\em Yellow}: $0.063-0.188$, {\em Yellow-Green}: $0.188-0.313$, {\em Green-Teal}: $0.313-0.438$, {\em Teal}: $0.438-0.563$, {\em Teal-Blue}: $0.563-0.688$, {\em Purple}: $0.688-0.813$, {\em Pink}: $0.813-0.938$\}. In the \cite{jern2013probabilistic} experiment, the hue of each experimenter-defined category was selected from one of six possible values, each of which falls into one of the color groups above. 
\jlanote{Note the two color bins we made that don't contain any of their values?}
By categorizing the participant-generated crystals likewise, we can obtain a broad measure of category contrast by determining the proportion of participant-generated crystals that fall into the same groups as the experimenter-defined categories: If contrast influences the hues of the generated categories, we should observe minimal overlap between in the color groupings.

@@

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern: The majority of participants in each condition ($n = 22$) generated categories possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the experimenter-defined categories. These results can be compared to the predictions of a Binomial model, which proposes that participants generate hues at random. That is, if hue selection is not systematic, the probability that any given example will lie in the same color group as an experimenter-defined category is given by a Binomial distribution with $p = 2/8=1/4$, as there were two experimenter-defined categories and eight possible color groups. Chi-square goodness-of-fit tests reveal that the observed distribution in each condition is highly inconsistent with the hues being chosen at random (all $\chi^2>200$, $p<0.001$): Participants tended to generate items that were perceptually distinct from the categories they had learned, and were less likely to generate hues possessed by members of the experimenter-defined categories.
\jlanote{need degrees of freedom for $\chi^2$ tests}
Re-analyzing the results of \cite{jern2013probabilistic} provides preliminary support that contrast plays a role in category generation. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our analysis suggests that generated categories tend to be distinct from {\em and} distributionally similar to what is already known. However, it is worth noting that our analysis is still limited: the color groups defined above are somewhat arbitrary, and it is not clear that our color grouping is consistent with psychological color boundaries. While we did obtain similar results using a variety of alternative groupings, the hue dimension used in the \cite{jern2013probabilistic} study does not lend itself straightforwardly to the the computation of similarities, and thus we cannot be certain of whether our coding accurately approximates the psychological space of the stimuli. This precludes exemplar modeling of their data as it is necessary to encode exemplars in psychological space to determine the similarity between exemplars. By consequence, although these results likely indicate that contrast exerts {\em some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

@@

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation inspired by factors influencing how PACKER generates new categories. Our experiments are based on \cite{jern2013probabilistic}'s paradigm: participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We then report formal analyses comparing PACKER's account of our results to that of the hierarchical sampling and copy-and-tweak models developed by \cite{jern2013probabilistic}.

@@

\nbcnote{can you think of other studies with size and color of squares as features?}

@@

In this way, one aspect of the analyses of the experimental results is a conceptual replication of the classic finding that generated categories tend to share distributional properties with known categories in the domain \citep[see][]{jern2013probabilistic,ward1994structured}. From these results, we can predict that, in each condition, participants should generate categories that are distributionally similar to the experimenter-defined category: In the Cluster condition, generated categories should be tightly clustered. In the Row condition, generated categories should vary more along the X-axis than the Y-axis. In XOR condition, generated categories should be widely distributed across both dimensions, and the two dimensions should be positively correlated.

Interestingly, the XOR condition also offers a dissociation between the roles of category contrast and the emulation of distributional structure: widely-distributed, positively-correlated categories would need to lie along the positive diagonal of the space (that is the only place they ``fit''), which is already occupied by the experimenter-defined category. Thus, if contrast plays a role, exemplars in the generated categories of participants in the XOR condition may not be positively correlated -- they may be negatively correlated instead. In this case, contrast and statistical regularities would interact, which would be inconsistent with the hierarchical sampling approach of creative cognition \citep{jern2013probabilistic}

@@

183 participants were recruited from Amazon Mechanical Turk. Participants were randomly assigned to one condition: 64 participants were assigned to the Cluster condition, 61 were assigned to the Row condition, and 58 were assigned to the XOR condition. Stimuli were squares varying in color (grayscale 9.8\%--90.2\%) and side length (3.0--5.8cm), see Figure \ref{fig:e1-conditions}. The assignment of perceptual features (color, size) to axes of the domain space (x, y), as well as the direction of variation along each axis (e.g., $dark \rightarrow light$ or $light \rightarrow dark$) was counterbalanced across participants.
\jlanote{we should probably mention the discrepancy between the number of participants per condition and explain why it isn't anything to worry about.}

@@

% In Experiment 1 we sought to extend our analysis of the \cite{jern2013probabilistic} data by evaluating the influence of category contrast on creative generation, given qualitatively different types of prior knowledge. We found strong evidence for effects of category contrast in each condition: Participants were more likely to generate stimuli that are more distant from (i.e., less similar to) members of a previously-learned category, and members of participant-generated categories tended to be more similar to one another than to members of previously-learned categories. We also partially replicated the classic finding that the distributional structure of generated categories reflects that of previously learned categories \citep{jern2013probabilistic,ward1994structured}: members of generated categories were more widely distributed along dimensions which were widely distributed in the experimenter-defined category.

@@

% Thus, the core results of Experiment 1 indicate that generated categories systematically differ from what can be expected based on prior work. The negative (or zero) correlations observed in the XOR condition suggests an interesting interaction between contrasting from a given category and emulating statistical properties. That is, the constraints on creative generation imposed by category contrast may not simply influence the {\em location} of generated categories, but also their distributional structure. In Experiment 2, we test this claim more systematically.

@@

To test whether category contrast influences the distributional structure of generated categories, we sought to identify conditions in which differences in the distributional structure of generated categories cannot be explained by the distributional structure of the experimenter-defined category. We created two new category types (depicted in Figure ~~\ref{fig:e2-conditions}) that possess an identical distributional structure (both are tight clusters of examples with no correlation between features), and only differ in their Y-axis position: the `Bottom' category lies in the bottom-center of the space, and the `Middle' category lies in the center. The distributional equality of these conditions is key to the design of the experiment: If the structure of previously learned categories were the only influence on the structure of generated categories, we should observe no difference between these two conditions with respect to distributional structure.

@@

To explore the aggregate distributional structure of the results, we computed the range of exemplars along each axis (X, Y), as well as the correlation between features. These data, shown in Figure \ref{fig:e2-statsboxes}, again demonstrate the degree of individual differences observed in our study. In each condition, we observed tightly clustered and widely distributed categories along each dimension, as well as positively, negatively, and uncorrelated categories.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: if contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical sampling model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical sampling model and PACKER is meant to emphasize the necessity of contrast and demonstrate it cannot be explained through the emulation of distributional structure: whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-category similarity), the hierarchical sampling model generates categories exclusively on the basis of knowledge of how existing classes are distributed. Our approach in this section is to first broadly evaluate and compare the quality of each model's account to our entire dataset (Experiments 1 and 2 combined), then analyze the ability for each model to explain individual differences in each experiment, and lastly we more describe the strengths and weakness of each model's account of category generation.

@@

Table \ref{table:global-model-fits} contains the results of this fitting procedure. Due to the uneven number of fitted parameters among the models, we compare the model fits using the Akaike Information Criterion \citep[AIC;][]{akaike1974new}, where smaller values correspond to better fits (discounted by the number of parameters). Table \ref{table:global-model-fits} contains the AIC values of each model's best fit, as well as the corresponding log-likelihood ($L$) and the best-fitting parameter values. These results reveal strong model differentiation: The PACKER model achieved far better fits compared to the copy-and-tweak and hierarchical sampling models, and copy-and-tweak performed somewhat better than the hierarchical sampling model. While PACKER's advantage may tentatively be attributed to the model's sensitivity to category contrast (this will be explored in detail below), the advantage shown by copy-and-tweak over the hierarchical sampling model may be attributed to its exemplar-based representation, as opposed to the prototype-based representation assumed in the hierarchical sampling model. As observed in Figures \ref{fig:e1-samples} and \ref{fig:e2-samples}, the generated categories we observed were often widely distributed, with no items near the category prototype. This aspect of the data is inconsistent with the multivariate normal distributions used to represent categories in the \cite{jern2013probabilistic} model, but can be handled easily using an exemplar-based approach.

A key distinction between PACKER and copy-and-tweak, as well as the hierarchical sampling model, is that, of the three models, only PACKER is capable of making strong predictions about the location of new category members when the target class is entirely novel (i.e., no member of the category has been observed). Under these circumstances, there are no examples to copy, and thus the copy-and-tweak model predicts that items are generated at random. Likewise, with no observations on which to condition the mean of the category distribution, the hierarchical sampling model also picks an item at random. Thus, it is possible that the failure of these models is simply due to their inability to explain each participant's first trial (generating the first item in the `Beta' category). We conducted an identical set of simulations as above, excluding this trial (leaving 915 responses in the dataset): Again, PACKER ($L = -3390$, $AIC = 6786$) achieved better fits than the copy-and-tweak ($L = -3579$, $AIC = 7162$) and hierarchical sampling ($L=-3612$, $AIC = 7232$) models. Finally, because copy-and-tweak is nested within PACKER, we can use a likelihood ratio test to compare the two models. PACKER explains the aggregate data significantly better than copy-and-tweak ($\chi^2(1) = 748, p < 0.001$ for all data and $\chi^2(1) = 378, p < 0.001$ excluding the first example), providing further evidence that category generation is better explained when contrast is considered.

\nbcnote{report something about robustness to model comparison metric (BIC and $AIC_c$).}
\jlanote{please -- i believe that they are derivable from log-likelihood, number of parameters, and number of data points}

@@

As in PACKER, changes to the weighting of the dimensions can also be used to produce row- and column-like categories in the copy-and-tweak and hierarchical sampling models. Indeed, as copy-and-tweak is simply a special case of the PACKER model, the attention weights operate exactly as described above. Within the hierarchical sampling model, the covariance matrix specifying the prior domain distribution, $\Sigma_0$, can be used to similar effect. This covariance matrix specifies the amount of variance assumed along each dimension (as well as the dimensional correlations) across the domain of categories. The covariance matrix for a newly generated category, $\Sigma_B$, is based on the assumed $\Sigma_0$ as well as the distributions of previously learned categories (see Appendix \ref{ap:hsampling-definition}). Thus, the importance of each feature can be coded into $\Sigma_0$ to alter the dimensional variance of generated categories.

However, a key limitation of both of the copy-and-tweak and hierarchical sampling models is the lack of interaction between the distributional structure of a category (in this case, a row-versus-column orientation) and the location of that category within the domain. While these models are {\em capable} of generating row and column categories, there is no mechanism in place to ensure generated categories will be distinct from what is already known. In the next subsection, we explore the interdependence between distributional structure and location in creative generation.

@@

Unlike the PACKER and copy-and-tweak models, the hierarchical sampling model's dimensional variances correspond to the assumed variance of generated categories along each dimension (rather than the inverse of the variance). Thus, a different transformation is appropriate for incorporating the weights computed in Equation \ref{eq:range-weight}. For the hierarchical sampling model, we computed the dimensional variances according to: $\lambda (1-w_k)d$, where $\lambda$ is a free parameter specifying the overall assumed variance of the domain, and $d$ is the number of dimensions. Under this approach, evenly distributed weights correspond to an assumed variance of $\lambda$. Likewise, larger values of $w$, which are produced when the generated category is tightly distributed along one dimension, correspond to smaller assumed variances.

\nbcnote{This works for 2D, but not 3D and beyond (as $1-w_k$ would not produce a valid weighting). Maybe we should make a note of that, or design something more general.}
\jlanote{Isn't it just $1-\sum_{j\neq k}{w_j}$?}

Each model was used to simulate each participant's generation independently, with the importance of each dimension set according to the participant's generated category. The other free parameters within each model were set as in Table \ref{table:global-model-fits}. Every participant's generation was simulated \textbf{NUMBER} of times, resulting in \textbf{NUMBER * PARTICIPANTS} categories generated by each model. For comparison with our behavioral results, we then computed the range difference gradient identically as with the behavioral data. The results are shown in Figure \ref{fig:range-diff-gradients}.

@@

The creative generation of novel concepts is an intriguing yet understudied topic in cognitive science. While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: How is it that people are able to create something {\em different} from what is already known? We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-category and between-category similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

@@

\subsection{Similarity and Contrast in Cognition}
The proposal that category contrast is a primary constraint in creative generation is, in some ways, entirely commonsense. To successful create something \textit{new}, it must be different from what is known. Beyond its role in creative generation, category contrast is also of fundamental importance in categorization more broadly. All other factors held constant, new categories are easier to learn if they are dissimilar to members of other categories, and knowledge of highly distinct categories is applied more accurately than that of ill-defined categories\footnote{\nbcnote{This is an uncontroversial claim but we should have some cites. I just can't think of a super appropriate one right now...}}. Likewise, the act of forming category representations affects similarity judgments about category members and nonmembers, with category members being viewed as more similar to one another than members of other categories \citep{goldstone1994influences,goldstone2001altering} .

%discussion of color categories and how they evolve. how across cultures, they seem to maximize within-category similarity and minimize between-category simialrity

@@

@@

\subsection{Limitations and Future Directions}
\jlanote{this is where the importance sampling connection will come in and will involve specifying the computational problem precisely. that hasn't been done yet for category generation and then perhaps to creativity more broadly.}

@@

Finally, although an overall goal of this work is to advance the study of creative thought by explaining generation through the empirical and quantitative approaches from research in categorization, these approaches are not easily aligned with many of the myriad alternative theoretical viewpoints in the study of creativity \citep[for a review see][]{kozbelt2010theories}, such as those based on free association \citep{mednick1962associative} and conceptual combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we reduced our focus by studying a highly complex behavior (creative generation) as it applies within a well-established domain (artificial category learning). Importantly, however, our findings do provide a clue about a more general, computational-level interpretation of creative generation. That is, in this paper we have made clear that contrast is a key constraint, as it is required in order to successful create something new. However, it cannot be the only constraint: It is not enough for a generated category to be distinct, as categories generated purely on the basis of contrast would not be distributionally coherent. Thus, a more general interpretation of creative generation would need to concern both the distinctness and coherence of what is created.

@@

The generation of new concepts and ideas is a highly interesting topic, but it is difficult to study in a controlled experimental environment. In this paper, we have provided such an examination of category generation as it applies within an artificial categorization categorization experiment. Extending the literature on creative cognition, our experiments provide a detailed picture of the role of category contrast in generation: People seek to create concepts that are distinct from what they already know, and the nature what is created can be influenced by what \textit{does not} yet exist. Our simulations with traditional exemplar models, as well as a hierarchical sampling model, provide strong support for the claim that category contrast is of fundamental importance, and, more generally, demonstrate that popular explanatory approaches from basic research in cognitive science can offer a precise, quantitative account of a behavior as complex as that of creative generation.

>>>>

@@

\section{Something Different: A Role For Contrast}

@@

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, it is assumed that learners in previous experiments were {\em successful} in creating new categories. Thus, effects of category contrast should be observable within the experimental results of these studies. To provide a test of the influence of category contrast in creative generation, we conducted a novel analysis of Experiment 3 from \cite{jern2013probabilistic}\footnote{Although \cite{jern2013probabilistic} reported four experiments, we focus on Experiment 3 because their first two experiments tested generation of items into known categories, and their fourth experiment was identical to Experiment 3 but with a restricted generation space.}.

@@

Although we cannot know precisely which colors were displayed or perceived, we can still analyze their results from a coarse perspective to see whether there is preliminary support for contrast. To do so, we binned all possible hues into one of eight uniformly-spaced color groups: \{{\em Red}: $0-0.063$, $0.938-1$, {\em Yellow}: $0.063-0.188$, {\em Yellow-Green}: $0.188-0.313$, {\em Green-Teal}: $0.313-0.438$, {\em Teal}: $0.438-0.563$, {\em Teal-Blue}: $0.563-0.688$, {\em Purple}: $0.688-0.813$, {\em Pink}: $0.813-0.938$\}. In the \cite{jern2013probabilistic} experiment, the hue of each experimenter-defined category was selected from one of six possible values, each of which falls into one of the color groups above (two color groups do not contain a possible hue for the experimenter-defined categories). By categorizing the participant-generated crystals likewise, we can obtain a broad measure of category contrast by determining the proportion of participant-generated crystals that fall into the same groups as the experimenter-defined categories: If contrast influences the hues of the generated categories, we should observe minimal overlap between in the color groupings.

@@

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern: The majority of participants in each condition ($n = 22$) generated categories possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the experimenter-defined categories. These results can be compared to the predictions of a Binomial model, which proposes that participants generate hues at random. That is, if hue selection is not systematic, the probability that any given example will lie in the same color group as an experimenter-defined category is given by a Binomial distribution with $p = 2/8=1/4$, as there were two experimenter-defined categories and eight possible color groups. Chi-square goodness-of-fit tests reveal that the observed distribution in each condition is highly inconsistent with the hues being chosen at random, all $\chi^2(6,N=22)>200$, $p<0.001$. Participants tended to generate items that were perceptually distinct from the categories they had learned, and were less likely to generate hues possessed by members of the experimenter-defined categories.

Re-analyzing the results of \cite{jern2013probabilistic} provides preliminary support that contrast plays a role in category generation. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our analysis suggests that generated categories tend to be distinct from {\em and} distributionally similar to what is already known. However, it is worth noting that our analysis is still limited: the color groups defined above are somewhat arbitrary, and it is not clear that our color grouping is consistent with psychological color boundaries. While we did obtain similar results using a variety of alternative groupings, the hue dimension used in the \cite{jern2013probabilistic} study does not lend itself straightforwardly to the the computation of similarities, and thus we cannot be certain of whether our coding accurately approximates the psychological space of the stimuli. This precludes exemplar modeling of their data as it is necessary to encode exemplars in psychological space to determine the similarity between exemplars. By consequence, although these results likely indicate that contrast exerts {\em some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

@@

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation inspired by factors influencing how PACKER generates new categories. Our experiments are based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward translation of the traditional artificial classification paradigm to the task of category generation: participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We then report formal analyses comparing PACKER's account of our results to that of the hierarchical sampling and copy-and-tweak models developed by \cite{jern2013probabilistic}.

@@

@@

In this way, our experimental results provide a conceptual replication of the classic finding that generated categories tend to share distributional properties with known categories in the domain \citep[see][]{jern2013probabilistic,ward1994structured}. From these results, we can predict that, in each condition, participants should generate categories that are distributionally similar to the experimenter-defined category: In the Cluster condition, generated categories should be tightly clustered. In the Row condition, generated categories should vary more along the X-axis than the Y-axis. In XOR condition, generated categories should be widely distributed across both dimensions, and the two dimensions should be positively correlated.

Interestingly, the XOR condition also offers a dissociation between the roles of category contrast and the emulation of distributional structure: widely-distributed, positively-correlated categories would need to lie along the positive diagonal of the space (that is the only place they ``fit''), which is already occupied by the experimenter-defined category. Thus, if contrast plays a role, exemplars in the generated categories of participants in the XOR condition may not be positively correlated -- they may be negatively correlated instead. In this case, contrast and statistical regularities would interact, which would be inconsistent with the leading approaches in conceptual generation \citep{jern2013probabilistic}

@@

183 participants were recruited from Amazon Mechanical Turk. Each participant was randomly assigned to one condition: 64 participants were assigned to the Cluster condition, 61 were assigned to the Row condition, and 58 were assigned to the XOR condition (sample sizes differ due to random assignment). Stimuli were squares varying in color (grayscale 9.8\%--90.2\%) and side length (3.0--5.8cm), see Figure \ref{fig:e1-conditions}. The assignment of perceptual features (color, size) to axes of the domain space (x, y), as well as the direction of variation along each axis (e.g., $dark \rightarrow light$ or $light \rightarrow dark$) was counterbalanced across participants.

@@

In Experiment 1 we sought to extend our analysis of the \cite{jern2013probabilistic} data by evaluating the influence of category contrast on creative generation, given qualitatively different types of prior knowledge. We found strong evidence for effects of category contrast in each condition: Participants were more likely to generate stimuli that are more distant from (i.e., less similar to) members of a previously-learned category, and members of participant-generated categories tended to be more similar to one another than to members of previously-learned categories. We also partially replicated the classic finding that the distributional structure of generated categories reflects that of previously learned categories \citep{jern2013probabilistic,ward1994structured}: members of generated categories were more widely distributed along dimensions which were widely distributed in the experimenter-defined category.

@@

Thus, the core results of Experiment 1 indicate that generated categories systematically differ from what can be expected based on prior work. The negative (or zero) correlations observed in the XOR condition suggests an interesting interaction between contrasting from a given category and emulating statistical properties. That is, the constraints on creative generation imposed by category contrast may not simply influence the {\em location} of generated categories, but also their distributional structure. In Experiment 2, we test this claim more systematically.

@@

To test whether category contrast influences the distributional structure of generated categories, we sought to identify conditions in which differences in the distributional structure of generated categories cannot be explained by the distributional structure of the experimenter-defined category. We created two new category types (depicted in Figure \ref{fig:e2-conditions}) that possess an identical distributional structure (both are tight clusters of examples with no correlation between features), and only differ in their Y-axis position: the `Bottom' category lies in the bottom-center of the space, and the `Middle' category lies in the center. The distributional equality of these conditions is key to the design of the experiment: If the structure of previously learned categories were the only influence on the structure of generated categories, we should observe no difference between these two conditions with respect to distributional structure.

@@

To explore the distributional structure of the generated categories, we computed the range of exemplars along each axis (X, Y), as well as the correlation between features. These data, shown in Figure \ref{fig:e2-statsboxes}, again demonstrate the degree of individual differences observed in our study. In each condition, we observed tightly clustered and widely distributed categories along each dimension, as well as positively, negatively, and uncorrelated categories.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: if contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical sampling model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical sampling model and PACKER is meant to emphasize the necessity of contrast and demonstrate that generation cannot be explained entirely through the emulation of distributional structure: whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-category similarity), the hierarchical sampling model generates categories exclusively on the basis of knowledge of how existing classes are distributed.

Our approach in this section is to first broadly evaluate and compare the quality of each model's account to our entire dataset (Experiments 1 and 2 combined), then analyze the ability for each model to explain individual differences in each experiment, and lastly we describe the strengths and weakness of each model's account of category generation.

@@

Table \ref{table:global-model-fits} contains the results of this fitting procedure. Due to the uneven number of fitted parameters among the models, we compare the model fits using the Akaike Information Criterion \citep[AIC;][]{akaike1974new}, where smaller values correspond to better fits (discounted by the number of parameters). Similar results were obtained with alternative model comparison metrics (e.g., BIC, \citealp{schwarz1978estimating}; $AIC_C$, \citealp{hurvich1989regression})

Table \ref{table:global-model-fits} contains the AIC values of each model's best fit, as well as the corresponding log-likelihood ($L$) and the best-fitting parameter values. These results reveal strong model differentiation: The PACKER model achieved far better fits compared to the copy-and-tweak and hierarchical sampling models, and copy-and-tweak performed somewhat better than the hierarchical sampling model. While PACKER's advantage may tentatively be attributed to the model's sensitivity to category contrast (this will be explored in detail below), the advantage shown by copy-and-tweak over the hierarchical sampling model may be attributed to its exemplar-based representation, as opposed to the prototype-based representation assumed in the hierarchical sampling model. As observed in Figures \ref{fig:e1-samples} and \ref{fig:e2-samples}, the generated categories we observed were often widely distributed, with no items near the category prototype. This aspect of the data is inconsistent with the multivariate normal distributions used to represent categories in the \cite{jern2013probabilistic} model, but can be handled easily using an exemplar-based approach.

A key distinction between PACKER and copy-and-tweak, as well as the hierarchical sampling model, is that, of the three models, only PACKER is capable of making strong predictions about the location of new category members when the target class is entirely novel (i.e., no member of the category has been observed). Under these circumstances, there are no examples to copy, and thus the copy-and-tweak model predicts that items are generated at random. Likewise, with no observations on which to condition the mean of the category distribution, the hierarchical sampling model also picks an item at random. Thus, it is possible that the failure of these models is simply due to their inability to explain each participant's first trial (generating the first item in the `Beta' category). We conducted an identical set of simulations as above, excluding this trial (leaving 915 responses in the dataset): Again, PACKER ($L = -3390$, $AIC = 6786$) achieved better fits than the copy-and-tweak ($L = -3579$, $AIC = 7162$) and hierarchical sampling ($L=-3612$, $AIC = 7232$) models.

Finally, because copy-and-tweak is nested within PACKER, we can use a likelihood ratio test to compare the two models. PACKER explains the aggregate data significantly better than copy-and-tweak ($\chi^2(1) = 749, p < 0.001$ for all data and $\chi^2(1) = 377, p < 0.001$ excluding the first example), providing further evidence that category generation is better explained when contrast is considered.

@@

As in PACKER, changes in the parameter settings of the copy-and-tweak model can also be used to produce different patterns of generation. Indeed, as copy-and-tweak is simply a special case of the PACKER model, the attention weights operate exactly as described above to produce row- and column-like categories. However, because the model is not influenced by category contrast, it is biased toward generating tightly clustered categories, as new items are always most likely to be generated near known examples of the target category. Thus, the lack of a contrast mechanism prevents the model from explaining why some individuals widely distribute their categories to the corners of the space.

Similar to the copy-and-tweak model, the hierarchical sampling model possesses no mechanism to account for category contrast, and the model is most likely to generate new items that are similar to known examples of the target category. Although this precludes an account of why individuals might assume a cluster versus corners profile, the covariance matrix specifying the model's prior domain distribution, $\Sigma_0$, can be used to explain the generation of row-like and column-like categories. This covariance matrix specifies the amount of variance assumed along each dimension (as well as the dimensional correlations) across the domain of categories. The covariance matrix for a newly generated category, $\Sigma_B$, is based on the assumed $\Sigma_0$ as well as the distributions of previously learned categories (see Appendix \ref{ap:hsampling-definition}). Thus, the importance of each feature can be coded into $\Sigma_0$ to alter the dimensional variance of generated categories.

As noted above, while the copy-and-tweak and hierarchical sampling models possess mechanisms to explain row- and column-like categories, they cannot easily explain why some individuals widely distribute their generated categories into disparate corners of the space. This, however, reveals a more general limitation: According to the copy-and-tweak and hierarchical sampling models, the distributional structure of generated categories is {\em independent} of their location within the domain. For example, although the copy-and-tweak or hierarchical sampling models can be parameterized to generate row- or column-like categories, there is no mechanism in place to ensure that what is generated will be distinct from what is already known. In the next subsection, we explore this prediction through an analysis of the interdependence between distributional structure and location in creative generation.

@@

Unlike the PACKER and copy-and-tweak models, the hierarchical sampling model's dimensional variances correspond to the assumed variance of generated categories along each dimension (rather than the inverse of the variance). Thus, a different transformation is appropriate for incorporating the weights computed in Equation \ref{eq:range-weight}. For the hierarchical sampling model, we computed the dimensional variances according to: $\lambda (1-w_k)d$, where $\lambda$ is a free parameter specifying the overall assumed variance of the domain, and $d$ is the number of dimensions\footnote{This calculation applies only in two-dimensional domains, where $w_2 = 1-w_1$.}. Under this approach, evenly distributed weights correspond to an assumed variance of $\lambda$. Likewise, larger values of $w$, which are produced when the generated category is tightly distributed along one dimension, correspond to smaller assumed variances.

Each model was used to simulate each participant's generation independently, with the importance of each dimension set according to the participant's generated category. The other free parameters within each model were set as in Table \ref{table:global-model-fits}. Every participant's generation was simulated 2,000 times; multiplied against 305 total participants, each model generated 610,000 categories in total. For comparison with our behavioral results, we then computed the range difference gradient identically as with the behavioral data. The results are shown in Figure \ref{fig:range-diff-gradients}.

@@

The creative generation of concepts is an greatly intriguing topic, but it is also of great complexity and can be difficult to study in a traditional laboratory setting. Creative generation is most typically studied using a creative cognition approach \citep{finke1992creative,smith1995creative}, where creative products \citep[such as drawings of alien plants and animals;][]{ward1994structured} are analyzed to obtain insight into the processes and representations involved. While this approach has yielded a great deal of information about the nature of creative generation, the commonly employed response modalities (e.g., hand-drawn pictures) made difficult the development of computational models to formally evaluate the proposed representations and processes. However, the generation of novel concepts is not altogether different from the types of behaviors studied in cognitive psychology: In particular, it can be viewed as an form of inference using one's knowledge about existing categories \citep{kemp2014taxonomy,kurtz2015human}. Taking advantage of this insight, in this paper we investigated the generation of new categories using the empirical and computational toolkit from the field of categorization.

While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: How is it that people are able to create something {\em different} from what is already known? We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-category and between-category similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

@@

\subsection{Similarity and Contrast in Cognition}
The proposal that category contrast is a primary constraint in creative generation is, in some ways, entirely commonsense. To successful create something \textit{new}, it must be different from what is known. Beyond its role in creative generation, category contrast is also of fundamental importance in categorization more broadly. All other factors held constant, new categories are easier to learn if they are dissimilar to members of other categories, and knowledge of highly distinct categories is applied more accurately than that of ill-defined categories \citep{ashby1994categorization,imai1965discriminability}. Likewise, basic-level categories \citep{rosch1976basic} are thought to be abstracted in order to maximize within-category similarity while minimizing between-category similarity. Finally, the act of forming category representations affects similarity judgments about category members and nonmembers, with category members being viewed as more similar to one another than members of other categories \citep{goldstone1994influences,goldstone2001altering} .

@@

\nbcnote{Something about analogy (similar relationally but different feature-ally); generated categories are alike in distributional structure but differently located in domain}

@@

\subsection{Limitations and Future Directions}

@@

Finally, although an overall goal of this work is to advance the study of creative thought by explaining generation through the empirical and quantitative approaches from research in categorization, these approaches are not easily aligned with many of the alternative theoretical viewpoints in the broader study of creativity \citep[for a review see][]{kozbelt2010theories}, such as those based on free association \citep{mednick1962associative} and conceptual combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we reduced our focus by studying a highly complex behavior (creative generation) as it applies within a well-established domain (artificial category learning).

@@

The generation of new concepts and ideas is a highly interesting topic, but it is difficult to study in a controlled experimental environment. In this paper, we have provided such an examination of category generation as it applies within an artificial categorization categorization experiment. Extending the literature on creative cognition, our experiments provide a detailed picture of the role of category contrast in generation: People seek to create concepts that are distinct from what they already know, and the nature of what is created can be influenced by what does not yet exist. Our simulations with traditional exemplar models, as well as a hierarchical sampling model, provide strong support for the claim that category contrast is of fundamental importance, and, more generally, demonstrate that popular explanatory approaches from basic research in cognitive science can offer a precise, quantitative account of a behavior as complex as that of creative generation.
